Symmetric Circuits for Rank Gates

### 2017-11-28 sections/symmetric-circuits.tex
28 Nov 2017

%AD: re-wrote the introductory paragraphs to be a bit more substantive.

@@

% Boolean circuits have long acted as an alternative model from which to study
% algorithmic complexity. A circuit is evaluated for some input binary sequence by
% recursively evaluating each gate in the circuit, starting from the input gates.
% Suppose then we in the process of evaluating the circuit $C$ and are at the
% stage of evaluating the gate $g$. We take the input sequence formed from the
% evaluations of the gates input to $g$ and applying the element of the basis that
% labels $g$ to that sequence. It follows that in order to evaluate $g$ we need to
% suppose some order on the evaluations of the input gates. In many cases this
% order is simply induced by an assumed order on the whole of the circuit; an
% assumption common in most parts of complexity theory, where the encoding of
% structures for computation induces such orders, but problematic in finite model
% theory.

% Circuit models have also been studied in finite model theory, often as a means
% for characterising the expressive power of logics. In particular, research has
% focused on symmetric circuits defined over bases of symmetric Boolean functions
% (see \cite{}). The fact that the basis consist of only symmetric functions is a
% crucial requirement, ensuring that the evaluation of a gate does not depend on
% an external order. However, in this paper we will need to consider circuits with
% gates that compute rank, a function that is not symmetric.

% In this section we extend this circuit model so as to allow for the inclusion of
% non-symmetric functions in the basis. We allow each gate $g$ to index its input
% gate in accordance with the basis element labeling $g$. We introduce the notion
% of a \emph{structured function}, a Boolean function whose input is indexed in
% accordance with a many-sorted vocabulary $\tau$. The input vector to a
% structured function can then be thought of as encoding a $\tau$-structure. In
% particular, we consider structured functions that compute properties of
% $\tau$-structures, which we call \emph{isomorphism-invariant} structured
% functions. We then define what it means for a circuit over a basis of
% isomorphism-invariant structured functions to be symmetric. This generalises the
% notion of a symmetric circuit. We also show that there are limitations on the
% power of symmetric circuits over bases of symmetric functions, justifying the
% importance of this generalisation.

@@

%AD - rewrote the paragraph to reduce the use of subscripts and superscripts.
Let $\tau = (R, S, \nu)$ be a many-sorted signature.
% with $R = \{R_1 \ldots,
% R_t\}$, $S = \{s_1, \ldots , s_p\}$, and $\nu(R_i) = (a^i_1, \ldots ,
% a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
% [r_i]$.
Let $D := \biguplus_{s \in S} D_{s} = \{(s,d) : d \in D_s\}$, be a
disjoint union of non-empty sets, indexed by $S$, i.e.\ $D$ is
potentially the universe of a $\tau$-structure. We call $D$ a \emph{$\tau$-sorted set}.
For each $P \in R$, let $P^D$ denote the trivial relation over the
many-sorted universe $D$, i.e.\ $P^D := D_{a_1} \times \ldots \times
D_{a_{r}}$, where $\nu(P) = (a_1, \ldots , a_{r)})$.   Let the \emph{index defined by the pair $(\tau, D)$} be
the set $\biguplus_{P\in R} P^{D}$.  We use this set to index the input to a Boolean
function, as such we abuse notation and denote the index defined by
$(\tau, D)$ by $\ind(\tau, D)$. We call $\tau$ the vocabulary and $D$ the universe of $\ind(\tau, D)$.

@@

We now define a \emph{$(\mathbb{B}, \tau)$-circuit} . This is a circuit that
computes a function on structures in the relational vocabulary $\tau$ and where
the gates are labelled by elements of the basis $\mathbb{B}$. Each element of
the basis is a $(\tau,D)$-structured function for some $\tau$ and $D$. Note
that, in the interests of keeping our definitions from getting excessively
complicated, we do not consider circuits computing functions over multi-sorted
vocabularies.
\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis and $\tau$ be a relational vocabulary, we define a
  \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$ computing a $q$-ary query $Q$ as a
  structure $\langle G, \Omega, \Sigma, \Lambda, L \rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ such that $\vert \Sigma^{-1} (0) \vert \leq 1$ and $\vert
    \Sigma^{-1} (1) \vert \leq 1$. Those gates mapped to $\tau \uplus \{0,1\}$
    are called input gates, with those mapped to $\tau$ called relational gates
    and those mapped $\{0,1\}$ called constant gates. Those gates mapped to
    $\mathbb{B}$ are called internal gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R_i})_{R_i \in
      R}$ such that $\Lambda_{R_i}$ maps each relational gate $g$ with $\Sigma
    (g) = R_i$ to the tuple $\Lambda_{R_i} (g) \in [n]^{r}$. When no ambiguity
    arises we write $\Lambda (g)$ for $\Lambda_{R_i} (g)$.
  \item $L$ associates with each internal gate $g$ a function $L(g):
    \ind(\Sigma(g)) \rightarrow G$ such that if we define a relation $W
    \subseteq G^{2}$ by $W(h_1,h_2)$ iff $h_2$ is an internal gate and $h_1$ is
    in the image of $L(h_2)$, then $(G, W)$ is a directed acyclic graph.
  \end{itemize}
\end{definition}

@@

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit. The $n$ here denotes that $C_n$ accepts structures of size $n$.
For a gate $g \in C_n$, we define the \emph{index} of $g$, denoted by $\ind(g)$,
as $\ind(\Sigma(g))$ and we let $H_g$ denote the set of all gates in the image
of $L(g)$. We also refer to the vocabulary and universe of $g$ to abbreviate
referring to the vocabulary and universe of $\Sigma(g)$.

@@

Let $\tau$ be a relational vocabulary, % = (R_1, \ldots , R_t)$
$\mathcal{A}$ a $\tau$-structure over an $n$-element universe $U$ and $\gamma
\in [n]^{\underline{U}}$.
% AD - need to define the notation [n]^{\underline{U}} in the background section
% -- DONE

Let $\gamma \mathcal{A}$ be the structure on universe $[n]$ formed by mapping
$U$ in accordance with $\gamma$. The evaluation of a $(\mathbb{B},
\tau)$-circuit $C_n$ computing a $q$-ary query $Q$ proceeds by recursively
evaluating the gates in the circuit. The evaluation of the gate $g$ for the
bijection $\gamma$ and input $\mathcal{A}$ is denoted by $C[\gamma
\mathcal{A}](g)$, and is given as follows:
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate such that $\Sigma (g)$ let $L^{\gamma}_g:
  \ind(g) \rightarrow \{0,1\}$ be defined by $L^{\gamma}_g(x) = C[\gamma
  \mathcal{A}](L(g)(x))$, for all $x \in \ind(g)$. Then $g$ evaluates to true
  iff $\Sigma(g) (L^{\gamma}_g) = 1$.
\end{enumerate}
We say that $C_n$ defines the $q$-ary query $Q \subseteq U^q$ under $\gamma$
where $\vec{a} \in Q$ iff $C_n[\gamma \mathcal{A}](\Omega (\gamma \vec{a})) =
1$.

@@

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. The function computed by $C_n$ is
  $\tau$-invariant iff $C_n$ is an invariant circuit.
\end{lem}
% AD - does this require a few lines of proof? To be added

@@

@@

@@

It follows that for any symmetric circuit $C_n$ there is a homomorphism $h$ that
maps $\sym_n$ to $\aut(C)$, taking any $\sigma \in \sym_n$ to an automorphism
that extends it. Such a homomorphism is injective so long as some element of
$[n]$ appears in a label of some input gate of $C$ (as then all elements appear
by symmetry)~\cite{AndersonD17} and therefore no two distinct elements of
$\sym_n$ agree on all input gates. In this paper we always assume that this is
the case, as otherwise all inputs are constant, and so the circuit just computes
a constant function. In order to develop a great many technical tools needed in
this paper we should like to be able to, without a loss of generality, assume
that that this homomorphism is in fact an isomorphism as well. In order to do so
we make use of the rigidity condition.

We define rigidity in terms of an equivalence relation on the gates of a
circuit. This recursively defined equivalence relation holds between two gates
if the circuit underneath these two gates are in some sense the same. Before we
define this relation we need to introduce some notation.

@@

We have already required that for each gate $g$ in a circuit must have an
injective labelling function. The following condition ensures that $L(g)/{\sim}$
is also an injection.

\begin{definition}
  Let $C_n$ be a circuit and $g$ be a gate in $C_n$. We say $g$ has \emph{unique
    labels} if $\vert H_g \vert = \vert H_g \setminus \equiv \vert $. We say
  that the circuit $C_n$ has \emph{unique labels} if every gate in $C_n$ has
  unique labels.
\end{definition}

We are also ready to define what it means for a circuit to be rigid.

\begin{definition}
  We say that a circuit $C_n$ is \emph{rigid} if every equivalence class of
  gates under the syntactic equivalence relation is a singleton.
\end{definition}

We prove in Proposition~\ref{} that there is a polynomial time algorithm that
takes in a symmetric circuit with unique labels and outputs an equivalent rigid
circuit. It is worth noting that the requirement that the circuit have unique
labels is very important (see Section \ref{} for a detailed discussion of this
point). Since we are interested in $P$-uniform families of circuits, this result
allows us to assume the rigidity of a circuit without a loss of generality.
Moreover, as we will show in Proposition \ref{}, if a circuit is rigid then
every $\sigma \in \sym_n$ has exactly one an extension to an automorphism of
$C_n$, and so the above defined homomorphism mapping $\sym_n$ to $\aut(C_n)$ is
an isomorphism. Since we may assume rigidity without a loss of generality this
property may also be assumed without a loss of generality. As such we abuse
notation and let $\sigma \in \sym_n$ also denote the induced automorphism of the
circuit.

@@

Let $\MB$ be the set of all structured functions that are symmetric or
matrix-invariant. Let $r, p \in \nats$, with $p$ prime, and let $D_1$ and $D_2$
be initial segments of the natural numbers. Then there is a matrix-invariant
structured function $F^{\rank}_{p,r}: \{0,1\}^{D_1 \times D_2}$, with universe
$D_1 \uplus D_2$, such that $F^{\rank}_{p,r}(M) = 1$ if, and only if, the matrix
$M \in \{0,1\}^{D_1 \times D_2}$ has rank at most $r$ when its entries are
understood as elements of the finite field $\mathbb{F}_p$. Let
$\mathbb{B}_\rank$ be the union of the set of all such functions with
$\mathbb{B}_{\maj}$.

\begin{definition}
  Let $\mathbb{B} \subseteq \MB$. We call a symmetric $(\mathbb{B},
  \tau)$-circuit with unique labels a \emph{symmetric matrix-circuit}.
\end{definition}

\begin{definition}
  Let $\mathbb{B} \subseteq \MB$. We call a symmetric $(\mathbb{B}_\rank,
  \tau)$-circuit with unique labels a \emph{symmetric rank-circuit}.
\end{definition}

>>>>

% AD: re-wrote the introductory paragraphs to be a bit more substantive.

@@

% Boolean circuits have long acted as an alternative model from which to study
% algorithmic complexity. A circuit is evaluated for some input binary sequence
% by recursively evaluating each gate in the circuit, starting from the input
% gates. Suppose then we in the process of evaluating the circuit $C$ and are at
% the stage of evaluating the gate $g$. We take the input sequence formed from
% the evaluations of the gates input to $g$ and applying the element of the
% basis that labels $g$ to that sequence. It follows that in order to evaluate
% $g$ we need to suppose some order on the evaluations of the input gates. In
% many cases this order is simply induced by an assumed order on the whole of
% the circuit; an assumption common in most parts of complexity theory, where
% the encoding of structures for computation induces such orders, but
% problematic in finite model theory.

% Circuit models have also been studied in finite model theory, often as a means
% for characterising the expressive power of logics. In particular, research has
% focused on symmetric circuits defined over bases of symmetric Boolean
% functions (see \cite{}). The fact that the basis consist of only symmetric
% functions is a crucial requirement, ensuring that the evaluation of a gate
% does not depend on an external order. However, in this paper we will need to
% consider circuits with gates that compute rank, a function that is not
% symmetric.

% In this section we extend this circuit model so as to allow for the inclusion
% of non-symmetric functions in the basis. We allow each gate $g$ to index its
% input gate in accordance with the basis element labeling $g$. We introduce the
% notion of a \emph{structured function}, a Boolean function whose input is
% indexed in accordance with a many-sorted vocabulary $\tau$. The input vector
% to a structured function can then be thought of as encoding a
% $\tau$-structure. In particular, we consider structured functions that compute
% properties of $\tau$-structures, which we call \emph{isomorphism-invariant}
% structured functions. We then define what it means for a circuit over a basis
% of isomorphism-invariant structured functions to be symmetric. This
% generalises the notion of a symmetric circuit. We also show that there are
% limitations on the power of symmetric circuits over bases of symmetric
% functions, justifying the importance of this generalisation.

@@

% AD - rewrote the paragraph to reduce the use of subscripts and superscripts.
Let $\tau = (R, S, \nu)$ be a many-sorted signature.
% with $R = \{R_1 \ldots, R_t\}$, $S = \{s_1, \ldots , s_p\}$, and $\nu(R_i) =
% (a^i_1, \ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for
% all $q \in [r_i]$.
Let $D := \biguplus_{s \in S} D_{s} = \{(s,d) : d \in D_s\}$, be a disjoint
union of non-empty sets, indexed by $S$, i.e.\ $D$ is potentially the universe
of a $\tau$-structure. We call $D$ a \emph{$\tau$-sorted set}. For each $P \in
R$, let $P^D$ denote the trivial relation over the many-sorted universe $D$,
i.e.\ $P^D := D_{a_1} \times \ldots \times D_{a_{r}}$, where $\nu(P) = (a_1,
\ldots , a_{r)})$. Let the \emph{index defined by the pair $(\tau, D)$} be the
set $\biguplus_{P\in R} P^{D}$. We use this set to index the input to a Boolean
function, as such we abuse notation and denote the index defined by $(\tau, D)$
by $\ind(\tau, D)$. We call $\tau$ the vocabulary and $D$ the universe of
$\ind(\tau, D)$.

@@

We now define a \emph{$(\mathbb{B}, \tau)$-circuit} . This is a circuit that
computes a function on structures in the relational vocabulary $\tau$ and where
the gates are labelled by elements of the basis $\mathbb{B}$. Each element of
the basis is a $(\tau,D)$-structured function for some $\tau$ and $D$. Note
that, in the interests of keeping our definitions from getting excessively
complicated, we do not consider circuits computing functions over multi-sorted
vocabularies.

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis and $\tau$ be a relational vocabulary, we define a
  \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$ computing a $q$-ary query $Q$ as a
  structure $\langle G, \Omega, \Sigma, \Lambda, L \rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ such that $\vert \Sigma^{-1} (0) \vert \leq 1$ and $\vert
    \Sigma^{-1} (1) \vert \leq 1$. Those gates mapped to $\tau \uplus \{0,1\}$
    are called input gates, with those mapped to $\tau$ called relational gates
    and those mapped $\{0,1\}$ called constant gates. Those gates mapped to
    $\mathbb{B}$ are called internal gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R_i})_{R_i \in
      R}$ such that $\Lambda_{R_i}$ maps each relational gate $g$ with $\Sigma
    (g) = R_i$ to the tuple $\Lambda_{R_i} (g) \in [n]^{r}$. When no ambiguity
    arises we write $\Lambda (g)$ for $\Lambda_{R_i} (g)$.
  \item $L$ associates with each internal gate $g$ a function $L(g):
    \ind(\Sigma(g)) \rightarrow G$ such that if we define a relation $W
    \subseteq G^{2}$ by $W(h_1,h_2)$ iff $h_2$ is an internal gate and $h_1$ is
    in the image of $L(h_2)$, then $(G, W)$ is a directed acyclic graph.
  \end{itemize}
\end{definition}

@@

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit. The $n$ is used to indicates that $C_n$ accepts structures of
size $n$. For a gate $g \in C_n$, we define the \emph{index} of $g$, denoted by
$\ind(g)$, as $\ind(\Sigma(g))$. We define the relation $W \subseteq G^2$ by
$W(h,g)$ if, and only if, $h \in \range(L(g))$. For each $g \in G$ we let
$W(\cdot, g) := \{h \in G : W(h,g)\}$, and let $W(g, \cdot) := \{h \in G :
W(g,h)\}$. We call $W(\cdot, g)$ the \emph{children} of $g$ and $W(g, \cdot)$
the \emph{parents} of $g$. We also abbreviate $W(g, \cdot)$ by $H_g$. We let the
vocabulary and universe of $g$ be the vocabulary and universe of $\Sigma(g)$,
respectively.

@@

Let $\tau$ be a relational vocabulary, $\mathcal{A}$ a $\tau$-structure over an
$n$-element universe $U$ and $\gamma \in [n]^{\underline{U}}$. Let $\gamma
\mathcal{A}$ be the structure on universe $[n]$ formed by mapping $U$ in
accordance with $\gamma$. The evaluation of a $(\mathbb{B}, \tau)$-circuit $C_n$
computing a $q$-ary query $Q$ proceeds by recursively evaluating the gates in
the circuit. The evaluation of the gate $g$ for the bijection $\gamma$ and input
$\mathcal{A}$ is denoted by $C[\gamma \mathcal{A}](g)$, and is given as follows
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate such that $\Sigma (g)$ let $L^{\gamma}_g:
  \ind(g) \rightarrow \{0,1\}$ be defined by $L^{\gamma}_g(x) = C[\gamma
  \mathcal{A}](L(g)(x))$, for all $x \in \ind(g)$. Then $g$ evaluates to true
  iff $\Sigma(g) (L^{\gamma}_g) = 1$.
\end{enumerate}
We say that $C_n$ defines the $q$-ary query $Q \subseteq U^q$ under $\gamma$
where $\vec{a} \in Q$ if, and only if, $C_n[\gamma \mathcal{A}](\Omega (\gamma
\vec{a})) = 1$.

@@

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. The function computed by $C_n$ is
  $\tau$-invariant if, and only if, $C_n$ is an invariant circuit.
\end{lem}
% AD - does this require a few lines of proof? GW - To be added. I have not done
% this yet.

@@

% It can be shown that $P$-uniform families of invariant circuits over the
% standard basis can decide exactly those languages in $\PT$\cite{}. In this
% paper, we are instead interested in circuits with broader symmetry properties.
% We first define the notion of automorphism for circuits.

We now define the notion of an automorphism of circuit.

@@

We are particularly interested in circuits that have the property that every
permutation in $\sym_n$ extends to an automorphism of the circuit.

@@

It follows that for any symmetric circuit $C_n$ there is a homomorphism $h$ that
maps $\sym_n$ to $\aut(C_n)$, taking any $\sigma \in \sym_n$ to an automorphism
that extends it. Suppose $C_n$ contains no relational gates of non-zero arity.
In that case $C_n$ computes a constant function. As such, in this paper we
always assume a circuit contains at least one relational gate with non-zero
arity. There there exists a relational gate in $C_n$ with some element of $[n]$
appearing in the tuple labelling that gate. By symmetry it follows that all
elements of $[n]$ appear in tuples labelling relational gates in $C_n$. Thus no
two distinct elements of $\sym_n$ agree on all input gates, and so the
homomorphism $h$ is injective.

If this homomorphism is also surjective then we have that each element of
$\sigma$ extends uniquely to an automorphism of the circuit. Since many of the
technical tools developed in this paper require that circuits have this
property, we restrict our attention to classes of circuits that have
\emph{unique extensions}.

\begin{definition}
  We say that a circuit $C_n$ has \emph{unique extensions} if for every $\sigma
  \in \sym_n$ there is at most one $\pi_{\sigma} \in \aut(C_n)$ such that
  $\pi_{\sigma}$ extends $\sigma$.
\end{definition}

Anderson and Dawar~\cite{AndersonD17} introduce the notion of a \emph{rigid}
circuit, and show that, for circuits defined over a basis of symmetric functions
there is a polynomial-time algorithm that takes as input a symmetric circuit and
outputs an equivalent rigid symmetric circuit. They also show that rigid
circuits have unique extensions, allowing them to conclude that $P$-uniform
families of symmetric circuits over bases of symmetric functions may be assumed
to be rigid (and hence have unique extensions) without a loss of generality.

While we establish an approximately analogous argument, unfortunately, for the
general $(\mathbb{B}, \tau)$-circuit, establishing an exactly analogous
condition is a non-trivial problem. We discuss this problem in more detail in
Section \label{sec:transparency-and-unique-extensions}.

We now introduce the notion of a circuit with \emph{unique labels}, and show
that if a circuit has unique labels then it has has unique extensions. Before we
do so, we define the \emph{syntactic equivalence} relation on the gates of a
circuit.

% In order to prove this result, Anderson and Dawar introduce the notion of a
% \emph{rigid} circuit, proving that symmetric circuits may be

% However, it is not obvious that it is possible to generalise this result. We
% show in Proposition \ref{}, that for the general $(\mathbb{B}, \tau)$-circuit,
% the problem of deciding if a given circuit has unique extensions is at least
% as hard as the graph isomorphism problem. As such, we instead define a the
% notion of a circuit having \emph{unique labels} and show that if a circuit has
% unique labels then we may assume it has unique extensions. We then explicitly
% restrict ourselves to circuits with unique labels.

@@

It is easy to show that determining if two gates in a given circuit compute the
same function is $NP$-complete. The syntactic-equivalence relation defines a
weaker relation between gates, instead identifying gates that have `hereditary
similar' circuits defining them. It is easy to show that, for a circuit $C_n$,
if $\sigma \in \sym_n$ and $\pi, \pi' \in \aut(C_n)$ extend $\sigma$, then for
every gate $g$ in $C_n$, $\pi (g)$ and $\pi'(g)$ are syntactically equivalent.

We now introduce the \emph{unique labels} condition.

\begin{definition}
  Let $C_n$ be a circuit and $g$ be a gate in $C_n$. We say $g$ has
  \emph{injective labels} if $L(g)$ is an injection. We say $g$ has \emph{unique
    labels} if $L(g)$ has injective labels and no two gates in $\range(L(g))$
  are syntactically equivalent. We say $C_n$ has \emph{injective labels} (resp.
  \emph{unique labels}) if every gate in $C_n$ has injective labels (resp.
  unique labels). We say $C_n$ is \emph{transparent} if every non-symmetric gate
  in $C_n$ has unique labels.
\end{definition}

Let $C_n$ be a circuit and $W_t$ be the transitive closure of the $W$ relation
on $C_n$. We note that a circuit $C_n$ may have an internal gate $g$ such that
for every output gate $g'$, $\neg W_t(g, g')$ (i.e. there is no path from $g$ to
any output gate). We call such a gate \emph{redundant}, as their inclusion in
the circuit can not possibly have any bearing on the function computed by the
circuit. It can be shown that given a circuit $C_n$ with redundant gates we may
construct a new circuit $C_n'$ by simply removing the redundant gates. Moreover,
this process preserves important circuit properties and can be done efficiently.
We state and prove this formally in Lemma \ref{lem:redudant-gates}.

% \begin{definition}
%   Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a circuit
%   computing a $q$-ary query. We say a gate $g \in G$ is \emph{redundant} if
%   there is no $\vec{a} \in [n]^q$ such that $W_t(g, \Omega (\vec{a}))$.
% \end{definition}

% We note that the definition of a circuit allows for the inclusion of gates
% that are never used when evaluating the circuit. That is, for a circuit $C_n$
% let $W_t$ be the transitive closure of $W$. It may be the case that there
% exists a gate $h$ in $C_n$ such that there is no output gate $g$ with $W_t(h,
% g)$.

\begin{lem}
  There exists an algorithm that takes as input a $(\mathbb{B}, \tau)$-circuit
  $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ computing a $q$-ary
  query and outputs a $(\mathbb{B}, \tau)$-circuit $C_n'$ such that (i) every
  gate in $C_n$ is in $C_n'$, (ii) $C_n$ and $C_n'$ compute the same query,
  (iii) no gate in $C_n'$ is redundant, (iv) if $C_n$ is symmetric then $C_n'$
  is symmetric, (v) if $C_n$ has unique labels then $C_n'$ has unique labels,
  and (vi) if $C_n$ is transparent then $C_n'$ is transparent. This algorithm
  runs in time polynomial in the size of the input circuit.
  \label{lem:redudant-gates}
\end{lem}
\begin{proof}
  We construct the sub-circuit $C_n'$ of $C_n$ by removing all redundant gates
  in $C_n$. Since $W_t$, the transitive closure of $W$, can be computed in time
  polynomial in the size of the circuit, this construction can be completed in
  time polynomial in the size of $C_n$.

  Conditions (i), (ii), (iii), (v) and (vi) follow obviously for $C_n'$. Suppose
  $C_n$ is symmetric and let $\sigma \in \sym_n$. Let $\pi$ be an automorphism
  of $C_n$ extending $\sigma$. Let $g$ be a gate in $C_n'$. Then there exists
  $\vec{a} \in [n]^q$ such that $W_t(g, \Omega(\vec{a}))$ in $C_n$. It can be
  shown that $W_t (\pi(g), \Omega(\sigma(\vec{a})))$. Thus $\pi (g)$ is in
  $C_n'$, and so restricting $\pi$ to the gates of $C_n'$ is an automorphism of
  $C_n'$ extending $\sigma$. It follows that $C_n'$ is symmetric.
\end{proof}

Given Lemma \ref{lem:redudant-gates}, we may assume, without a loss of
generality, that a circuit has no redundant gates. In this paper we make this
assumption of every circuit unless stated otherwise.

We now show that unique labels are sufficient for unique extensions.

\begin{prop}
  Let $C_n := \langle G, \Omega, \Sigma, \Lambda, L\rangle$ be a $\mathbb{B},
  \tau)$-circuit. If $C_n$ has unique labels then $C_n$ has unique extensions.
  \label{prop:unique-labels-unique-extensions}
\end{prop}
% \begin{prop}
%   Let $C_n := \langle G, \Omega, \Sigma, \Lambda, L\rangle$ be an injective
%   $(\mathbb{B}, \tau)$-circuit. Then $C_n$ has unique labels if, and only if,
%   $C_n$ has unique extensions.
% \end{prop}
\begin{proof}
  Suppose $C_n$ has unique labels. Let $\pi_{\sigma}, \pi_{\sigma}' \in
  \aut(C_n)$ be automorphisms extending $\sigma \in \sym_n$. We now prove that
  $\pi_\sigma = \pi_{\sigma}'$. It is sufficient to show that $\pi :=
  \pi_{\sigma}'\pi^{-1}_\sigma$ is the trivial automorphism on the circuit. It
  is easy to see that $\pi$ extends the trivial permutation $e \in \sym_n$. Thus
  $\pi$ acts trivially on input gates and output gates in the circuit, as $\pi$
  acts as $e$ on relational gates and for any output gate $g$ there exists
  $\vec{a} \in [n]^q$ such that $\pi (g) = \pi (\Omega(\vec{a})) = \Omega(e \dot
  \vec{a}) = \Omega(\vec{a}) = g$.

  It can be shown by induction that, since $\pi$ extends the trivial
  permutation, for all gates $g \in G$, $\pi(g)$ is syntactically equivalent to
  $g$. Let $W_t$ be the transitive closure of the relation $W$ on $C_n$. Let $g
  \in G$ and suppose $\pi (g) = g$. Then $\pi H_{g} = H_g$ and, since $g$ has
  unique labels, it follows that $\pi$ acts trivially on $H_g$ (i.e. for all $h
  \in H_{g}$, $\pi(h) = h$). It follows by induction that for all $h \in G$ if
  $W_t(h, g)$ then $\pi (h) = h$. Since the circuit contains no redundant gates,
  it follows that $G = \bigcup_{\vec{a} \in [n]^q} \{g \in G : W_t (g,
  \Omega(\vec{a}))\}$. Thus, since $\pi$ acts trivially on output gates, we have
  that for all $g \in G$, $\pi (g) = g$. The result follows.

  % Suppose $C_n$ does have not have unique labels. For every such gate $g \in
  % G$ let $D(g)$ be the distance from $g$ to the set of output gates. From the
  % hypothesis there exists $g \in G$ such that $h, h' \in H_g$ and $h \equiv
  % h'$. Let $g$ the gate in the circuit without unique labels and with minimal
  % $D(g)$. Then

  % to the set of output gates Order the set of all such gates by their distance
  % to the set of output gates, and let $g$ be the minimal gate in this set.

  % such that it is of minimal distance to the set of output gates. Let $e \in
  % \sym_n$ be the trivial permutation and $\pi \in \aut(C_n)$ be the unique
  % extension of $e$. We have that $\pi (g) = g$, as if $\pi (g) \neq g$
\end{proof}

We also show that a transparent circuit can be converted into an equivalent
circuit with unique labels in time polynomial in the size of the circuit. Since
we are working with $P$-uniform families of transparent circuits, this fact
allows us to assume, without a loss of generality, that transparent circuits
have unique labels. We prove this formally in Proposition
\ref{prop:transparent-unique}.

% \begin{definition}
%   We say that a circuit $C_n$ is \emph{rigid} if every equivalence class of
%   gates under the syntactic equivalence relation is a singleton.
% \end{definition}
% \

% We prove in Proposition~\ref{} that there is a polynomial time algorithm that
% takes in a symmetric circuit with unique labels and outputs an equivalent
% rigid circuit. It is worth noting that the requirement that the circuit have
% unique labels is very important (see Section \ref{} for a detailed discussion
% of this point). Since we are interested in $P$-uniform families of circuits,
% this result allows us to assume the rigidity of a circuit without a loss of
% generality. Moreover, as we will show in Proposition \ref{}, if a circuit is
% rigid then every $\sigma \in \sym_n$ has exactly one an extension to an
% automorphism of $C_n$, and so the above defined homomorphism mapping $\sym_n$
% to $\aut(C_n)$ is an isomorphism. Since we may assume rigidity without a loss
% of generality this property may also be assumed without a loss of generality.
% As such we abuse notation and let $\sigma \in \sym_n$ also denote the induced
% automorphism of the circuit.

@@

% Let $\MB$ be the set of all structured functions that are symmetric or
% matrix-invariant.

In many contexts, circuits are defined over a basis of symbols, usually the
usual logical symbols $\land, \neg, \lor$ and perhaps $\maj$ (for majority).
When evaluating a circuit, a gate labelled by a particular symbol is evaluated
using a Boolean function appropriate to the particular symbol. We have instead
defined a basis to be a set of Boolean functions. However, when reasoning about
circuits or dealing with them as inputs to algorithms we often want to identify
a function in the basis with a particular symbol. We now formally define a few
natural bases used in this paper, assigning to each function in these bases an
appropriate symbol.

Let $a \in \nats$, let $\land[a] : \{0,1\}^{[a]} \rightarrow \{0,1\}$ be the
structured function computing the $\AND$ operation on $a$ inputs. Let $\lor[a] :
\{0,1\}^{[a]} \rightarrow \{0,1\}$ be the structured function computing the
$\OR$ operation on $a$ inputs. Let $\maj [a] : \{0,1\}^{[a]} \rightarrow
\{0,1\}$ be the structured function computing the the majority function on $a$
inputs. We abuse notation and let $\neg : \{0,1\} \rightarrow \{0,1\}$ also
denote the structured function computing negation on a single input. In each
case we use the name of the function above as a symbol denoting that function.

Let the \emph{standard basis} $\BS$ be the basis containing $\neg$ and all
functions $\land[a]$ and $\lor[a]$ for every $a \in \nats$. Let the
\emph{majority basis }$\BM := \BS \cup \{\maj[a] : a \in \nats\}$. We say a
basis $\BB$ is \emph{symmetric} if every function in $\mathbb{B}$ is a symmetric
Boolean function.

Let $a, b, r, p \in \nats$, with $p$ prime. Let $\rank^r_p[a,b] : \{0,1\}^{[a]
  \times [b]} \rightarrow \{0,1\}$ be a matrix-invariant structured with
universe $[a] \uplus [b]$, such that $\rank^r_p[a,b](M) = 1$ if, and only if,
the matrix $M \in \{0,1\}^{[a] \times [b]}$ has rank at most $r$ when its
entries are understood as elements of the finite field $\mathbb{F}_p$. We again
identify the function $\rank^{r}_p[a,b]$ with the symbol $\rank^{r}_p[a,b]$.

Let the \emph{\rank basis} $\RB := \BM \cup \{\rank^{r}_p[a,b] : a, b, r, p \in
\nats, \text{ $p$ prime}\}$. We say a basis $\BB$ is \emph{matrix-symmetric} if
every function in $\BB$ is symmetric or matrix-symmetric.

% Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit and let $g$ be a gate in $C_n$. We
% say $g$ is an \emph{\AND-gate} if $g$ is an internal gate labelled a function
% of the . We say $g$ is a a \emph{majority gate}

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit and let $g \in G$. We say a gate $g \in G$ is a \emph{symmetric
  gate} if $\Sigma(g)$ is a symmetric function. We say a gate $g \in G$ is a
\emph{matrix-symmetric gate} if $\Sigma(g)$ is a matrix-symmetric structured
function. We say that $g$ is an \emph{$\AND$ gate} if $\Sigma(g) := \land [a]$
for some $a \in \nats$. Similarly we define \emph{$\OR$- gates}, \emph{$\NOT$
  gates}, \emph{$MAJ$ gates} and \emph{$\rank$ gates}.
